Subglacial and subaerial fluvial sediment transport capacity respond differently to water discharge variations

### 2024-06-27 Delaneyetal2024.tex
reworking

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
% AGUJournalTemplate.tex: this template file is for articles formatted with LaTeX
% 
% This file includes commands and instructions
% given in the order necessary to produce a final output that will
% satisfy AGU requirements, including customized APA reference formatting.
% 
% You may copy this file and give it your
% article name, and enter your text.
% 
% guidelines and troubleshooting are here:

%% To submit your paper:
\documentclass[draft]{agujournal2019}
\usepackage{url} %this package should fix any errors with URLs in refs.
\usepackage{lineno}
\usepackage{placeins}
\usepackage{amsmath}
\usepackage{makecell}

\usepackage[inline]{trackchanges} %for better track changes. finalnew option will compile document with changes incorporated.
\usepackage{soul}
\linenumbers
%% make line numbers work with amsmath
% https://tex.stackexchange.com/questions/461186/how-to-use-lineno-with-amsmath-align
\usepackage{lineno}   %% <- no mathlines option
\usepackage{amsmath}  %% <- after lineno
\usepackage{etoolbox} %% <- for \cspreto, \csappto

%% Patch 'normal' math environments:
\newcommand*\linenomathpatch[1]{%
  \cspreto{#1}{\linenomath}%
  \cspreto{#1*}{\linenomath}%
  \csappto{end#1}{\endlinenomath}%
  \csappto{end#1*}{\endlinenomath}%
}

\linenomathpatch{equation}
\linenomathpatch{gather}
\linenomathpatch{multline}
\linenomathpatch{align}
\linenomathpatch{alignat}
\linenomathpatch{flalign}

%%%%%%% 
% As of 2018 we recommend use of the TrackChanges package to mark revisions.
% The trackchanges package adds five new LaTeX commands:
% 
% \note[editor]{The note}
% \annote[editor]{Text to annotate}{The note}
% \add[editor]{Text to add}
% \remove[editor]{Text to remove}
% \change[editor]{Text to remove}{Text to add}
% 
% complete documentation is here: http://trackchanges.sourceforge.net/
%%%%%%%

\draftfalse

\newcommand{\aref}[1]{\textbf{Reference #1}}
\newcommand{\TODO}[1]{\textbf{TODO: \color{red}#1}}
\newcommand{\ian}[1]{{\textbf{\color{blue}Ian says:} \color{blue} #1} }
\newcommand{\mauro}[1]{{\textbf{\color{green}Mauro says:} \color{green} #1} }
\newcommand{\alpine}{\textit{ALPINE}}
\newcommand{\icesheet}{\textit{ICESHEET}}
\newcommand{\m}{$\,\mathrm{m}$\,}
\newcommand{\cm}{$\,\mathrm{cm}$\,}
\newcommand{\mma}{$\,\mathrm{mm  \, a^{-1}}$\,}
\newcommand{\mmma}{$\,\mathrm{m^3\, a^{-1}}$\,}
\newcommand{\mmms}{$\,\mathrm{m^3\, s^{-1}}$\,}
\newcommand{\unit}[1]{$\mathrm{#1}$}

\journalname{Geophysical Research Letters}

@@

\authors{Ian Delaney\affil{1},  Andrew J. Tedstone\affil{2},   Mauro A. Werder\affil{3,4},  Daniel Farinotti\affil{3,4} }

\affiliation{1}{Institut des dynamiques de la surface terrestre (IDYST), Universit\'{e} de Lausanne, B\^{a}timent G\'{e}opolis, 1015 Lausanne, Switzerland}
\affiliation{2}{Department of Geosciences, University of Fribourg, Ch. du Musée 1700, Fribourg, Switzerland}
\affiliation{3}{Laboratory of Hydraulics, Hydrology and Glaciology (VAW), ETH-Z\"urich, H\"onggerbergring 26, 8093 Z\"urich, Switzerland}
\affiliation{4}{Swiss Federal Institute for Forest, Snow and Landscape Research (WSL) Z\"uricherstrasse 111, 8903 Birmensdorf, Switzerland}

\correspondingauthor{Ian Delaney}{ianarburua.delaney@unil.ch}

\begin{keypoints}
\item Rapid water discharge fluctuations in subglacial channels are accommodated by water velocity, not channel size as in subaerial systems.
\item Greater variability in water velocity causes sediment transport capacity to vary more in subglacial channels than in subaerial channels.
\item Significant subglacial sediment transport capacity can persist across a large range of water discharges, impacting  the timing of peak events.
\end{keypoints}

\begin{abstract}
  Sediment transport capacity in both subaerial and subglacial channels depends on the shear stress exerted across the channel bottom, which varies with water velocity and channel width.
  In subaerial channels, water discharge variations are accommodated  by flow depth and width changes, along with water velocity.
  However, in subglacial channels,  water discharge changes mainly result in velocity variations, as the channel geometry evolves slowly (over days).
  Here, hydrographs from an Alpine glacier and the Greenland Ice Sheet are applied to subglacial and subaerial hydraulics models, revealing greater variability in the inferred sediment transport capacity in subglacial than in subaerial channels.
  Results also show that high subglacial sediment transport rates and water velocities can occur across a wide range of water discharges.
  These findings are discussed in the context of understanding sediment discharge from glaciers with different hydro-climatic forcings and assumptions of channel shape.
\end{abstract}

\section*{Plain Language Summary} % 200 words
Sediment export from glaciers impacts downstream ecosystems and communities and has changed in response to climate warming.
Sediment export depends on two main factors: sediment availability, and the capacity of the hydrological system to transport that sediment.
This capacity largely depends on the width of the channel and the water's velocity.
In subaerial channels, like rivers, water velocity, and channel width both respond rapidly to water discharge.
However, subglacial channels are pressurized by the ice above, so water flows through a subglacial pipe that evolves in size only slowly.
Variations in water discharge and thus also sediment transport capacity are therefore mainly accommodated by changes in the water velocity.
Models of these processes show that sediment transport capacity is highly variable in subglacial channels, and a wide range of water discharges can result in similar sediment transport capacities.
In subaerial channels, by contrast, sediment transport capacity is less variable, and variations depend solely on water discharge, not the channel size.
These findings may impact the interpretation of sediment transport records from glacierized catchments, complicating the link between glacier melt and sediment discharge.

\section{Introduction}
\label{sect:intro}

Changes in glacier dynamics and hydrology  have prompted numerous  recent studies of  sediment transport processes in cold regions \cite<e.g.>{li2022,vergara2022,zhang2022}.
Increases in sediment transport  have been observed in Greenland \cite{bendixen2017}, the European Alps \cite{costa2017}, the Himalayas \cite{li2021}, and the Andes \cite{vergara2022}.
In some regions, increased water discharge and glacier melt have been interpreted to yield greater sediment transport capacity \cite{bendixen2017,costa2017,li2021}.
To explain observed changes to sediment transport in glacierized catchments requires examining the processes controlling sediment discharge and its variations with water discharge  \cite<e.g.>{riihimaki2005,swift2005}.

Periods of millennia or longer, processes such as glacier abrasion and quarrying sculpt landscapes and create sediment to be transported fluvially from under glaciers \cite<c.f.>{hallet1979,iverson2012,ugelvig2018}.
Pressurized subglacial water can transport this sediment from underneath glaciers \cite{walder1994,creyts2013,beaud2018,delaney2019}, if it is reachable by the water.

In a transport-limited regime, sediment discharge is controlled by sediment transport capacity, defined as the amount of sediment the water can carry.
In both subglacial and subaerial channels, sediment transport capacity depends on the shear stress between water and the sediment it flows  over \cite{shields1936,meyer1948,engelund1967} along with the width of the channel bottom $w$ over which sediment mobilizes.
The shear stress $\tau$ responds to the velocity of water $v$ flowing through the channel so that
\begin{linenomath*}
  \begin{equation}
    \label{eq:tau}
    \tau \propto v^2.
  \end{equation}
\end{linenomath*}
% 
Following mass conservation, the mean velocity of the water flowing through a  channel is
\begin{linenomath*}
  \begin{equation}
    \label{eq:v}
    v = \frac{Q}{S},
  \end{equation}
\end{linenomath*}
where $Q$ is water discharge,  and $S$ is the channel's wetted area.

In subaerial channels, operating with open channel flow, the wetted area $S$ of the channel evolves with changing water discharge $Q$, by changing both the channel's width and the water depth \cite{leopold1953}.
The change in water depth results in a proportional increase in water velocity, and  the shear stress $\tau$ increases according to Equation~\ref{eq:tau}.

The response of water velocity to changing water discharge in subglacial channels differs from subaerial ones.
The size of subglacial channels is controlled by the opposing processes of channel opening by frictional heating of water flow on the one hand versus creep closure by ice flow on the other \cite{rothlisberger1972}.
As a result, the subglacial channel size only evolves relatively slowly over days, whereas water discharge can vary more quickly over hours \cite<e.g.>{iken1986,andrews2014,nanni2020}.
Therefore, on short periods, subglacial water flow behaves more like pipe flow, and changes in water discharge $Q$ are mainly accommodated by changing water velocity $v$ \cite<Equation~\ref{eq:v} and Figure~\ref{fig:cartoon}; >{alley1997}.

\begin{figure}[h]
  \centering
  \includegraphics[width=0.8\linewidth]{Fig1.pdf}
  \caption{Sketch for the different responses of subglacial and subaerial channels to increased water discharge over short time-scales.
    Arrow length denotes water velocity magnitudes in the subglacial (subaerial) $v_g$ ($v_f$) channels.
    $S_g$ ($S_f$) represents the wetted area in subglacial (subaerial) channels.
    The subglacial channel width $w_g$ remains unchanged, while the subaerial channel width $w_f$ evolves with water discharge.
    Subglacial (subaerial) shear stress $\tau_g$ ($\tau_f$) is responsible for the mobilization of sediment.
  }
  \label{fig:cartoon}
\end{figure}

Sediment mobilization in subaerial and subglacial channels, therefore, responds differently to changing water discharge.
These differences are implicitly included in a range of models quantifying sediment transport in both subglacial and subaerial channels  \cite<e.g.>{walder1994,alley1997,tucker1997,creyts2013,delaney2019,hewitt2019,wickert2019}.
Additionally, the variable response of sediment transport capacity to water discharge in the two systems may affect the interpretation of sediment transport records in glacial systems \cite<e.g.>{muller1968,richards2003,swift2005,ganti2016,pitlick2021}.
Yet, little work has examined the differing relationship between variations in water discharge and sediment discharge in subglacial and subaerial channels \cite{alley1997}.

This manuscript evaluates the relationship between sediment transport capacity and water discharge in subglacial channels and compares it to subaerial ones.
Models of subglacial and subaerial hydraulics are used to quantify the respective water velocity and channel morphology.
The models are applied to proglacial hydrological records from an Alpine glacier in Switzerland (Fieschergletscher) and  a land-terminating glacier in Greenland (Leverett Glacier).
The  lumped nature of the models isolates the relationship between water discharge and sediment transport capacity in subglacial and subaerial systems, independent of the upstream drainage network and sediment access.
Outputs demonstrate differences in the relationships between  water discharge, channel geometry, and water velocity in the subglacial and subaerial channels.
The manuscript then uses the demonstrated differences between the two channel types to discuss the implications for interpreting the processes driving variability in sediment transport records and their relationship with hydrology from glacierized catchments.

@@

Water discharge data is used from Alpine and ice sheet settings collected downstream of the glacier, assuming no water storage in the proglacial area.
The Alpine site (labeled \alpine{}) is  Fieschergletscher in the Swiss Alps ($46^\circ\,29'\,07''$ N, $8^\circ\,08'\,3''$ E).
The water discharge data used here was collected at a $1$\unit{min} interval from May 24, 2014, to October 10, 2014 \cite{felix2022}.

The Leverett Glacier in Greenland (labeled \icesheet{}) serves as the ice sheet setting.
Water discharge was measured roughly $2$\unit{km} downstream from the terminus ($67^\circ\,03'\,5''$ N, $50^\circ\,12'\,59''$ W), at a $5$\,\unit{min} time interval from May 28, 2012 to August 8, 2012 \citeA{tedstone2013}.

Note that we omit analysis of suspended sediment discharge records due to the role of sediment supply in capturing variations in sediment discharge, in addition to sediment transport capacity \cite<e.g.>{delaney2024}.

\section{Methods}
\label{sect:meth}
The two models described below (Sections ~\ref{sect:sub_mode}~and~\ref{sect:fluv}) represent relationships amongst water discharge, water velocity, and channel geometry in both subaerial and subglacial channels (Table \ref{table:vpm}).
Both models use the measured discharge to calculate water velocity, shear stress, and width-integrated shear stress, upon which both suspended sediment and bedload transport depend \cite<Figure \ref{fig:cartoon}; >{shields1936}.
Our evaluation in terms of shear stress omits the selection of a sediment transport relationship and a grain-size parameter \cite<e.g.>{shields1936,meyer1948}.

@@

To evaluate the shear stress of water flowing across sediments underneath a glacier, the subglacial channel model accounts for the channel geometry and the velocity of the flowing water.
To do this, we use a lumped hydraulics model from \citeA{werder2010b}, itself based upon  \citeA{clarke1996}.

Here, it is assumed that the water is transported through a subglacial channel \cite<Figure~\ref{fig:cartoon}; >{rothlisberger1972} beneath a glacier with channel length $l$, with a flat bed and a mean thickness of $h_{ice}$.
The channel  size grows from melt due to frictional heating from water flow and closes due to the creep  of the ice.
The formulation here does not consider the englacial storage of water.
The evolution of subglacial channel size $S_g$ is given as
\begin{linenomath*}
  \begin{equation}
    \label{eq:dS_dt}
    \frac{\partial S_g}{\partial t} = C_1 \frac{Q \Delta h}{l} - C_2 \left(h_{o}-\frac{\Delta h}{2}\right)^n\,S_g,
  \end{equation}
\end{linenomath*}
\noindent where $t$ is time, $C_1= (1-\rho_wc_pc_t)\,\frac{\rho_wg}{\rho_iL}$ and $C_2=2A(\frac{\rho_wg}{n})^n$ are constants (complete values given in Table \ref{table:vpm}), $g$ is the acceleration due to gravity, $Q$ is water discharge, $\Delta h$ is the hydraulic potential change over $l$, $h_{o}= \frac{\rho_i}{\rho_w} h_{ice}$ is the mean ice overburden pressure expressed in meter water equivalent ($\rho_w$ is density of water; $\rho_i$ is density of ice), and $n$ is Glen's n \cite{glen1955} (usually $n=3$).
The first term on the right side of the equation represents the opening of the channel through frictional heating, while the following term represents the creep closure of the channel from ice deformation.

Following the Darcy-Weisbach equation, the head drop $\Delta h$ is
\begin{linenomath*}
  \begin{equation}
    \label{eq:dh}
    \Delta h \,  = l \,\frac{1}{2g} \,f_i\,\frac{v_{g}^{2}}{D_h},
  \end{equation}
\end{linenomath*}
\noindent where $f_r$ is a friction factor, $D_h$ is the hydraulic diameter, $l$ is the channel length, and $v_g=\frac{Q}{S_g}$ is the water flow speed.
% 
The hydraulic diameter $D_h$ is converted to wetted area $S_g$ with
\begin{equation}
  \label{eq:Dh2S}
  S_g= \frac{D_h^2}{2}\,\frac{(\frac{\beta}{2}+\sin \frac{\beta}{2})^2}{\beta - \sin \beta},
\end{equation}
where $\beta$ is the central angle of the circular segment that comprises the channel (the so-called Hooke angle, \citeA{hooke1990}). Note that $\beta =\pi$ corresponds to a semi-circular channel and smaller values of $\beta$ result in shallow, wide channels.
This completes the subglacial hydraulic model which is described by the state variables $S_g$ and $\Delta h$.

The shear stress, $\tau_g$, between the water and the channel bed is determined through the Darcy-Weisbach formulation
\begin{equation}
  \label{eq:tau_g}
  \tau_g=\frac{1}{8}\,f_r\,\rho_w\,v_g^2,
\end{equation}
% 
where $v_g = \frac{Q}{S_g}$ is the water velocity.
% 
The width of the channel floor $w_g$ is represented as
\begin{equation}
  \label{eq:dh2wc}
  w_g = 2  \sin \frac{\beta}{2} \sqrt{\frac{2\, S_g}{\beta -\sin \beta}}.
\end{equation}
% 
This value is used to establish the integrated shear stress across the channel $w_g\tau_g$.

@@

\section{Results}
\begin{figure}[h]
  \centering
  \includegraphics[width=0.9\linewidth]{Fig2.png}
  \caption{Model outputs from simulations using the hydrographs in panels a and e for the scenarios \alpine{} (a-d) and \icesheet{} (e-h).
    Black (gray) lines in a and e represent the subglacial channel size (water discharge).
    Blue (orange) lines represent outputs from the subglacial (subaerial) channel.
    Data are shown at $15$\unit{min} intervals.
    Arrows show examples where variables peak in the subglacial channel before the subaerial one.
    Insets in f-h show the  peak melt event denoted by the shaded area in panels e--h, with an arbitrary y-axis.
    % Note that the x and y axes are different for \alpine{} and \icesheet{}.
  }
  \label{fig:model_outs}
\end{figure}

In the first experiment defined in Section~\ref{sect:imp}, model outputs exhibit different seasonal evolutions and peaks.
Water velocity and shear stress represent the potential for sediment mobilization, and width-integrated shear stress represents a proxy for the total sediment transport capacity across the channel bed (Figure~\ref{fig:model_outs}).
Additionally, the subglacial model outputs are generally larger than their subaerial equivalents, thus greater sediment transport capacity would be expected in subglacial channels compared to subaerial ones.

>>>>

%% Copernicus Publications Manuscript Preparation Template for LaTeX Submissions
%% ---------------------------------
%% This template should be used for copernicus.cls
%% The class file and some style files are bundled in the Copernicus Latex Package, which can be downloaded from the different journal webpages.
%% For further assistance please contact Copernicus Publications at: production@copernicus.org
%% https://publications.copernicus.org/for_authors/manuscript_preparation.html

%% Please use the following documentclass and journal abbreviations for preprints and final revised papers.

%% 2-column papers and preprints
\documentclass[esurf, manuscript]{copernicus}

\usepackage{makecell}
\usepackage{float}
\usepackage[section]{placeins}

\newcommand{\aref}[1]{\textbf{Reference #1}}
\newcommand{\afig}[1]{\textbf{Figure #1}}
\newcommand{\TODO}[1]{\textbf{TODO: \color{red}#1}}
\newcommand{\ian}[1]{{\textbf{\color{blue}Ian sayz:} \color{blue} #1} }
\newcommand{\mauro}[1]{{\textbf{\color{green}Mauro says:} \color{green} #1} }
\newcommand{\alpine}{\textit{ALPINE}}
\newcommand{\icesheet}{\textit{ICESHEET}}
\newcommand{\m}{$\,\mathrm{m}$\,}
\newcommand{\cm}{$\,\mathrm{cm}$\,}
\newcommand{\mma}{$\,\mathrm{mm  \, a^{-1}}$\,}
\newcommand{\mmma}{$\,\mathrm{m^3\, a^{-1}}$\,}
\newcommand{\mmms}{$\,\mathrm{m^3\, s^{-1}}$\,}
%\newcommand{\unit}[1]{$\mathrm{#1}$}

@@

% \Author[affil]{given_name}{surname}

\Author[1][IanArburua.Delaney@unil.ch]{Ian}{Delaney} %% correspondence author
\Author[2]{Andrew J.}{Tedstone}
\Author[3,4]{Mauro A.}{Werder}
\Author[3,4]{Daniel}{Farinotti}

\affil[1]{Institut des dynamiques de la surface terrestre (IDYST), Universit\'{e} de Lausanne, B\^{a}timent G\'{e}opolis, 1015 Lausanne, Switzerland}
\affil[2]{Department of Geosciences, University of Fribourg, Ch. du Musée 1700, Fribourg, Switzerland}
\affil[3]{Laboratory of Hydraulics, Hydrology and Glaciology (VAW), ETH-Z\"urich, H\"onggerbergring 26, 8093 Z\"urich, Switzerland}
\affil[4]{Swiss Federal Institute for Forest, Snow and Landscape Research (WSL) Z\"uricherstrasse 111, 8903 Birmensdorf, Switzerland}
%% The [] brackets identify the author with the corresponding affiliation. 1, 2, 3, etc. should be inserted.

%% If an author is deceased, please mark the respective author name(s) with a dagger, e.g. "\Author[2,$\dag$]{Anton}{Smith}", and add a further "\affil[$\dag$]{deceased, 1 July 2019}".

%% If authors contributed equally, please mark the respective author names with an asterisk, e.g. "\Author[2,*]{Anton}{Smith}" and "\Author[3,*]{Bradley}{Miller}" and add a further affiliation: "\affil[*]{These authors contributed equally to this work.}".

\runningtitle{Subglacial sediment transport capacity}

\runningauthor{Delaney et al.}
\received{}
\pubdiscuss{} %% only important for two-stage journals
\revised{}
\accepted{}
\published{}

%% These dates will be inserted by Copernicus Publications during the typesetting process.

\firstpage{1}

\maketitle

\begin{abstract}
  Sediment transport capacity in both subaerial and subglacial channels depends on the shear stress exerted across the channel bottom, which varies with water velocity and channel width.
  In subaerial channels, water discharge variations are accommodated  by flow depth and width changes, along with water velocity.
  However, in subglacial channels,  water discharge changes mainly result in velocity variations, as the channel geometry evolves slowly (over days).
  Here, we present relationships between water discharge and sediment transport under several channel types.
  They demonstrate that sediment transport is highly non-linear with respect to water discharge with fixed subglacial channel size, and in fact can very less that subaerial flow if water discharge changes slowly enough to be in equilibrium with subglacial channel size.
  We then apply hydrographs from an Alpine glacier and the Greenland Ice Sheet to subglacial and subaerial hydraulics models, revealing greater variability in the inferred sediment transport capacity in subglacial than in subaerial channels, as subglacial channels evolve in size.
  This variability stems from the slowly evolving channel size that forces water velocity to accommodate variations in water discharge.
  Furthermore, the changing channel size means that sediment transport capacity peaks prior to the maximum water discharge.
  Results also show that high subglacial sediment transport rates and water velocities can occur across a wide range of water discharges, meaning that in a transport-limited system there could be a variable relationship between sediment and water discharge.
  These findings are discussed in the context of understanding sediment discharge from glaciers with different hydro-climatic forcings and assumptions of channel shape.
  We also examine the implications for interpreting records of sediment discharge in glacierized catchments
\end{abstract}

\introduction  %% \introduction[modified heading if necessary]

Changes in glacier dynamics and hydrology  have prompted numerous  recent studies of  sediment transport processes in cold regions \citep[e.g.][]{li2022,vergara2022,zhang2022}.
Increases in sediment transport  have been observed in Greenland \citep{bendixen2017}, the European Alps \citep{costa2017}, the Himalayas \citep{li2021}, and the Andes \citep{vergara2022}.
To explain observed changes to sediment transport in glacierized catchments requires examining the processes controlling sediment discharge and its variations with water discharge  \citep[e.g.][]{riihimaki2005,swift2005}.

Periods of millennia or longer, processes such as glacier abrasion and quarrying sculpt landscapes and create sediment to be transported fluvially from under glaciers \citep[c.f.][]{hallet1979,iverson2012,ugelvig2018}.
Pressurized subglacial water can transport this sediment from underneath glaciers \citep{walder1994,creyts2013,beaud2018,delaney2019}, if it is reachable by the water.

In a transport-limited regime, sediment discharge is controlled by sediment transport capacity, defined as the amount of sediment the water can carry.
In both subglacial and subaerial channels, sediment transport capacity depends on the shear stress between water and the sediment it flows  over \citep{shields1936,meyer1948,engelund1967} along with the width of the channel bottom $w$ over which sediment mobilizes.
The shear stress $\tau$ responds to the velocity of water $v$ flowing through the channel so that
\begin{linenomath*}
  \begin{equation}
    \label{eq:tau}
    \tau \propto v^2.
  \end{equation}
\end{linenomath*}
% 
Following mass conservation, the mean velocity of the water flowing through a  channel is
\begin{linenomath*}
  \begin{equation}
    \label{eq:v}
    v = \frac{Q}{S},
  \end{equation}
\end{linenomath*}
where $Q$ is water discharge,  and $S$ is the channel's wetted area.

In subaerial channels, operating with open channel flow, the wetted area $S$ of the channel evolves with changing water discharge $Q$, by changing both the channel's width and the water depth \citep{leopold1953}.
The change in water depth results in a proportional increase in water velocity, and  the shear stress $\tau$ increases according to Equation~\ref{eq:tau}.

The response of water velocity to changing water discharge in subglacial channels differs from subaerial ones.
The size of subglacial channels is controlled by the opposing processes of channel opening by frictional heating of water flow on the one hand versus creep closure by ice flow on the other \citep{rothlisberger1972}.
As a result, the subglacial channel size only evolves relatively slowly over days, whereas water discharge can vary more quickly over hours \citep[e.g.][]{iken1986,andrews2014,nanni2020}.
Therefore, on short periods, subglacial water flow behaves more like pipe flow, and changes in water discharge $Q$ are mainly accommodated by changing water velocity $v$ \citep[Equation~\ref{eq:v} and Figure~\ref{fig:cartoon}; ][]{alley1997}.

\begin{figure}[hbt!]
  \centering
  \includegraphics[width=0.6\linewidth]{Fig1.pdf}
  \caption{Sketch for the different responses of subglacial and subaerial channels to increased water discharge over short time-scales.
    Arrow length denotes water velocity magnitudes in the subglacial (subaerial) $v_g$ ($v_f$) channels.
    $S_g$ ($S_f$) represents the wetted area in subglacial (subaerial) channels.
    The subglacial channel width $w_g$ remains unchanged, while the subaerial channel width $w_f$ evolves with water discharge.
    Subglacial (subaerial) shear stress $\tau_g$ ($\tau_f$) is responsible for the mobilization of sediment.
  }
  \label{fig:cartoon}
\end{figure}

Sediment mobilization in subaerial and subglacial channels, therefore, responds differently to changing water discharge.
These differences are implicitly included in a range of models quantifying sediment transport in both subglacial and subaerial channels  \citep[e.g.][]{walder1994,alley1997,tucker1997,creyts2013,beaud2018,delaney2019,hewitt2019,wickert2019}.
To date, explicit parameterizations of subglacial sediment transport capacity with respect to water discharge has assumed fixed channel size and demonstrates a very non-linear response in sediment transport capacity to water discharge \citep{alley1997}.
The modeling frameworks examining subglacial sediment transport with evolving channel size minimally discuss the impact of different hydrological forcing on sediment transport capacity with the aim of understanding the impact of hydro-climatic conditions on subglacial sediment dynamics \citep{creyts2013,beaud2018,delaney2019,hewitt2019}.
Yet, sediment discharge capacity controls the mobilization and deposition of sediment, making it one of the most fundamental aspects of subglacial sediment evacuation, especially over short timescales.

Despite the physical differences between the subaerial and subglacial systems, water discharge and sediment export are often compared in contemporary glacial environments \citep[e.g.][]{willis1996,pearce2003,richards2003,swift2005,chu2009,tedstone2012,chu2012,overeem2017,delaney2018,swift2021,lu2022,andresen2024}.
These comparisons are natural given that the majority of sediment evacuated from glaciers over contemporary seasonal timescales is done so fluvially \citep[e.g.][]{collins1989, riihimaki2005,delaney2024}.
In subaerial systems, water discharge and sediment transport capacity for a grain size are closely linked \citep[e.g. ][]{schmidt2006,pitlick2021} \ian{check citiation}.
However, the potential for a highly variable response in sediment transport capacity in subglacial sediment transport may affect the interpretation of sediment transport records in glacial systems \citep[e.g.][]{ganti2016,mancini2023}, if properties associate with subaerial systems are assumed.
This relationship should be clarified as sediment dynamics change in glacierized regions along with changing hydrology, increased variability of water discharge \citep{lane2019} and extreme glacier melt  \citep[e.g.][]{cremona2023,overeem2015} that may drive high magnitude sediment transport events.

This manuscript as two objectives: 1) evaluate the variability of subglacial sediment transport capacity with evolving subglacial channels size and different hydrographs and 2) establish if subseasonal variations in water discharge can be covary with sediment transport capacity in subglacial systems.
First, to reenforce the highly variable behavior of subglacial sediment transport capacity,  we present algebric formulations of the sediment transport capacity response to water discharge in subaerial, pipe-flow and equilibrium R-channel conditions, extending the relationships presented in \citet{alley1997}.
We then use numerical models to examine  sediment transport capacity in subglacial channels that evolve in size and subaerial channels.
They are then applied to proglacial hydrological records from an Alpine glacier in Switzerland (Fieschergletscher) and  a land-terminating glacier in Greenland (Leverett Glacier).
We run models with an ensemble of parameters to evaluate the sediment transport capacity variability in subglacial channels compared to subaerial ones.
Lastly, model outputs demonstrate the particular processes that can drive variability in sediment discharge capacity in subglacial systems.
Together, these findings indicate differences in the relationships between  water discharge, channel geometry, and water velocity in the subglacial and subaerial channels.
The manuscript then uses the identified differences between the subaerial and subglacial channel types to discuss the implications for interpreting the processes driving variability in sediment transport records and the impact of these differences on sediment export from glaciers.

@@

Water discharge data is used from Alpine and ice sheet settings collected downstream of the glacier, assuming no water storage in the proglacial area.
The Alpine site (labeled \alpine{}) is  Fieschergletscher in the Swiss Alps ($46^\circ\,29'\,07''$ N, $8^\circ\,08'\,3''$ E).
The water discharge data used here was collected at a $1$\unit{min} interval from May 24, 2014, to October 10, 2014 \citep[Figure~\ref{fig:Qw}\,a][]{felix2022}.

The Leverett Glacier in Greenland (labeled \icesheet{}) serves as the ice sheet setting.
Water discharge was measured roughly $2$\unit{km} downstream from the terminus ($67^\circ\,03'\,5''$ N, $50^\circ\,12'\,59''$ W), at a $5$\,\unit{min} time interval from May 28, 2012 to August 8, 2012 \citet[][Figure~\ref{fig:Qw}\,b]{tedstone2013}.

Note that we omit analysis of suspended sediment discharge records due to the role of sediment supply in capturing variations in sediment discharge, in addition to sediment transport capacity \citep[e.g.][]{delaney2019}.

\begin{figure}[hbt!]
  \centering
  \includegraphics[width=0.8\linewidth]{Qw_plot.pdf}
  \caption{Water discharge from Fieschergletcher (\alpine) and Leverett Glacier (\icesheet).
    Note differences in variability and quantity of water discharge over the season.
  }
  \label{fig:Qw}
\end{figure}

\section{Methods}
\label{sect:meth}
The two models described below (Sections ~\ref{sect:sub_mode}~and~\ref{sect:fluv}) represent relationships amongst water discharge, water velocity, and channel geometry in both subaerial and subglacial channels (Table~\ref{table:vpm}).
Both models use the measured discharge to calculate water velocity, shear stress, and width-integrated shear stress, upon which both suspended sediment and bedload transport depend \citep[Figure \ref{fig:cartoon}; ][]{shields1936}.
Our evaluation in terms of shear stress omits the selection of a sediment transport relationship and a grain-size parameter \citep[e.g.][]{shields1936,meyer1948}.
Through this choice we also negate sorting processes

@@

To evaluate the shear stress of water flowing across sediments underneath a glacier, the subglacial channel model accounts for the channel geometry and the water's velocity.
To do this, we use a lumped hydraulics model from \citet{werder2010b}, itself based upon  \citet{clarke1996}.

Here, it is assumed that the water is transported through a subglacial channel \citep[Figure~\ref{fig:cartoon}; ][]{rothlisberger1972} beneath a glacier with channel length $l$, with a flat bed and a mean thickness of $h_{ice}$.
The channel  size grows from melt due to frictional heating from water flow and closes due to the creep  of the ice.
The formulation here does not consider the englacial storage of water.
The evolution of subglacial channel size $S_g$ is given as
\begin{linenomath*}
  \begin{equation}
    \label{eq:dS_dt}
    \frac{\partial S_g}{\partial t} = C_1 \frac{Q \Delta h}{l} - C_2 \left(h_{o}-\frac{\Delta h}{2}\right)^n\,S_g,
  \end{equation}
\end{linenomath*}
\noindent where $t$ is time, $C_1= (1-\rho_wc_pc_t)\,\frac{\rho_wg}{\rho_iL}$ and $C_2=2A(\frac{\rho_wg}{n})^n$ are constants (complete values given in Table~\ref{table:vpm}), $g$ is the acceleration due to gravity, $Q$ is water discharge, $\Delta h$ is the hydraulic potential change over $l$, $h_{o}= \frac{\rho_i}{\rho_w} h_{ice}$ is the mean ice overburden pressure expressed in meter water equivalent ($\rho_w$ is density of water; $\rho_i$ is density of ice), and $n$ is Glen's n \citep{glen1955} (usually $n=3$).
The first term on the equation's right side represents the opening of the channel through frictional heating, while the following term represents the creep closure of the channel from ice deformation.

Following the Darcy-Weisbach equation, the head drop $\Delta h$ is
\begin{linenomath*}
  \begin{equation}
    \label{eq:dh}
    \Delta h \,  = l \,\frac{1}{2g} \,f_i\,\frac{v_{g}^{2}}{D_h},
  \end{equation}
\end{linenomath*}
\noindent where $f_r$ is a friction factor, $D_h$ is the hydraulic diameter, $l$ is the channel length, and $v_g=\frac{Q}{S_g}$ is the water flow speed.
% 
The hydraulic diameter $D_h$ is converted to wetted area $S_g$ with
\begin{equation}
  \label{eq:Dh2S}
  S_g= \frac{D_h^2}{2}\,\frac{(\frac{\beta}{2}+\sin \frac{\beta}{2})^2}{\beta - \sin \beta},
\end{equation}
where $\beta$ is the central angle of the circular segment that comprises the channel (the Hooke angle, \citet{hooke1990}). $\beta =\pi$ corresponds to a semi-circular channel and smaller values of $\beta$ result in shallow, wide channels.
This completes the subglacial hydraulic model which is described by the state variables $S_g$ and $\Delta h$.

The shear stress, $\tau_g$, between the water and the channel bed results from the Darcy-Weisbach formulation
\begin{equation}
  \label{eq:tau_g}
  \tau_g=\frac{1}{8}\,f_r\,\rho_w\,v_g^2,
\end{equation}
% 
where $v_g = \frac{Q}{S_g}$ is the water velocity.
% 
The width of the channel floor $w_g$ is represented as
\begin{equation}
  \label{eq:dh2wc}
  w_g = 2  \sin \frac{\beta}{2} \sqrt{\frac{2\, S_g}{\beta -\sin \beta}}.
\end{equation}
% 
This value is used to establish the integrated shear stress across the channel $w_g\tau_g$.

@@

In the second experiment, the models are  applied to a reference test case for each glacier.
These experiments assumes  a subglacial channel with $\beta=\frac{\pi}{6}$ and a subaerial channel with $\alpha = \frac{1}{3}$ and slope of $0.02$ (Table~\ref{table:vpm}).
Both friction factors $f_r$ and $f_f$ are tuned so that reasonable water velocities \citep[$\sim\,1.6\,$\unit{m}\,\unit{s}$^{-1}$][]{werder2010b,chandler2013} occur for both \alpine{} and \icesheet{}.
%
To quantify the link between sediment transport capacity and water discharge, model outputs are compared to water discharge from the glaciers using Spearman rank correlation, which accounts for the ordering of values, but not their magnitude.
Rank correlation reduces the impact of the non-linear response of sediment transport capacity to the hydrological forcing.

\section{Results}
We first present formulations of sediment transport capacity in several channel types to build upon the analysis in \citet{alley1997}.
We then apply the models above (Section~\ref{sect:meth}) in two experiments by examining three model outputs.
Water velocity and shear stress represent the potential for sediment mobilization, and width-integrated shear stress represents a proxy for the total sediment transport capacity across the channel bed (Figure~\ref{fig:model_outs}).

\subsection{Sensitivity of sediment transport in different drainage regimes}
\label{sect:scaling}
